---
title: "SEWS_HexGrid"
author: "Cesar A. Murad R."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load required libraries

library(terra)
library(raster)
library(sf)
library(dplyr)
library(stringr)
library(rgee)
library(ggplot2)
library(future.apply)
library(tools)
library(spatialwarningsGis)
library(spatialwarnings)
library(future)

```

### Landsat Composite Merger

This script merges raster from GEE exports to generate one composite per time period.

```{r}

# Location of exported GEE composites
rasters_folder <- "D:/Documents/QuebecTaigaStudy/Composites/SensitivityRobustness/CompositesVS" 
##setwd(composites_folder)

year_folders <- list.dirs(rasters_folder, recursive = FALSE) # Get list of year subfolders (non-recursive)

## Double Check if TIF files inside folders match the year folder

files <- list.files(rasters_folder, pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)

files_df <- data.frame(file = basename(files), folder_year = str_extract(basename(dirname(files)), "\\d{4}"),  file_year = str_extract(basename(files), "\\d{4}"), full_path = files) %>%
  mutate(match = folder_year == file_year)


mismatches <- files_df %>% dplyr::filter(!match)

if (nrow(mismatches) == 0) {
  message("✅ All files match their folder year.")
} else {
  message("⚠️ Some files do not match their folder year:")
  print(mismatches[, c("folder_year", "file_year", "file")])
}
```

Loop through each folder to merge all TIF files within the time period (same folder) into a single composite.

```{r}

for (yr_folder in year_folders) {
  
  t0    <- Sys.time()
  yr    <- basename(yr_folder)
  files <- list.files(yr_folder, pattern = "\\.tif$", full.names = TRUE)

  # Build the virtual mosaic
  vrt_file <- file.path(tempdir(), paste0("Composite_", yr, ".vrt"))
  v <- vrt(files, filename = vrt_file, overwrite = TRUE)

  # Materialize to a real GeoTIFF
  out_file <- file.path(rasters_folder, paste0("Composite_", yr, "_merged.tif"))
  writeRaster(v, out_file, overwrite = TRUE,
              gdal = c("TILED=YES", "BIGTIFF=IF_SAFER"))
  terra::tmpFiles(current = TRUE, remove = TRUE) # remove temp files

  message("Composite ", yr, ": ", length(files), " tiles, ",
          round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 2), " secs")
}
```

Build raster stack from generated mosaics

```{r}

merged_path <- "D:/Documents/QuebecTaigaStudy/Composites/LandsatComposites/FullBandsCompositesClean"
merged_files <- list.files(merged_path, pattern = "_merged\\.tif$", full.names = TRUE)

# Sort files by year if needed
# merged_files <- merged_files[order(merged_files)]

# Load into one multi-layer SpatRaster
merged_stack <- rast(merged_files)

```

### Calculate Vegetation Indices (VIs)

```{r}

# FUNCTION: compute supported indices from a named list of SpatRaster bands
compute_indices_from_bands <- function(bands, indices) {
  # bands: named list with elements blue, green, red, nir, swir1, swir2 as SpatRast layers (or NULL if missing)
  
  clean_band <- function(b) if (!is.null(b)) { b[b < 0] <- NA; return(b) } else NULL
  bands <- lapply(bands, clean_band)
  
  # convenience short-names
  b <- bands$blue;
  g <- bands$green;
  r <- bands$red;
  n <- bands$nir;
  s1 <- bands$swir1;
  s2 <- bands$swir2
  
  # indices: character vector of indices to compute
  out_layers <- list()
  
  for (idx in indices) {
    idx_upper <- toupper(idx)
    if (idx_upper == "NDVI") {
      # (NIR - RED) / (NIR + RED)
      if (is.null(n) || is.null(r)) stop("NDVI requires 'nir' and 'red' bands.")
      out_layers[["NDVI"]] <- (n - r) / (n + r)
    } else if (idx_upper == "EVI") {
      # 2.5 * (NIR - RED) / (NIR + 6*RED - 7.5*BLUE + 1)
      if (is.null(n) || is.null(r) || is.null(b)) stop("EVI requires 'nir', 'red', and 'blue' bands.")
      out_layers[["EVI"]] <- 2.5 * (n - r) / (n + 6 * r - 7.5 * b + 1 )
    } else if (idx_upper == "NDWI") {
      # (NIR - SWIR1) / (NIR + SWIR1)
      if (is.null(n) || is.null(s1)) stop("NDWI requires 'nir' and 'swir1' bands.")
      out_layers[["NDWI"]] <- (n - s1) / (n + s1)
    } else if (idx_upper == "SAVI") {
      # ((NIR - RED) * (1 + L)) / (NIR + RED + L) where L = 0.5
      if (is.null(n) || is.null(r)) stop("SAVI requires 'nir' and 'red' bands.")
      out_layers[["SAVI"]] <- ((n - r) * (1 + 0.5)) / (n + r + 0.5)
    } else if (idx_upper == "NBR") {
      # (NIR - SWIR2)/(NIR + SWIR2) - Normalized Burn Ratio
      if (is.null(n) || is.null(s2)) stop("NBR requires 'nir' and 'swir2' bands.")
      out_layers[["NBR"]] <- (n - s2) / (n + s2)
    } else if (idx_upper == "NDMI") {
      # (NIR - SWIR1)/(NIR + SWIR1) sometimes called NDMI or NBR2
      if (is.null(n) || is.null(s1)) stop("NDMI requires 'nir' and 'swir1' bands.")
      out_layers[["NDMI"]] <- (n - s1) / (n + s1)
    } else if (idx_upper == "LAI") {
      # 3.618*EVI - 0.118
      if (is.null(n) || is.null(r) || is.null(b)) stop("LAI requires 'nir', 'red', and 'blue' bands.")
      out_layers[["LAI"]] <- (3.618 * (2.5 * (n - r) / (n + 6 * r - 7.5 * b + 1 ))) - 0.118
    } else {
      stop("Index '", idx, "' is not implemented. Add it to compute_indices_from_bands() if desired.")
    }
  }
  return(out_layers)
}  
```

Compute VI from Landsat composites.

```{r}
## Main loop to calculate VIs from Landsat composites

indices_list <- c("NDVI","EVI","NDWI","NDMI","SAVI","NBR", "LAI")
VIndex <- indices_list[4] # 1 = NDVI - 2 = EVI - 3 = NDWI - 4 = NDMI - 5 = SAVI - 6 = NBR - 7 = LAI

composites_folder <- paste0("D:/Documents/QuebecTaigaStudy/Composites/",VIndex, "/Clean")  # output folder

if (dir.exists(composites_folder) == TRUE) {
  message(paste0("⚠️ Composite folder to store ",VIndex," composites already exists."))
} else {
  dir.create(composites_folder, showWarnings = FALSE, recursive = TRUE)
  message(paste0("✅ Composite folder for ",VIndex," composites successfully created.")) 
}

# Which indices to compute (any subset of supported indices below)
indices_to_compute <- c(VIndex)  

# Band mapping: set these to the band numbers of the merged multiband TIFF file
# e.g. common Landsat ordering in merged files: blue=1, green=2, red=3, nir=4, swir1=5, swir2=6
band_map <- list(blue = 1, green = 2, red = 3, nir = 4, swir1 = 5, swir2 = 6)

VIs_files <- character(0) # To store NDVI file paths

#merged_files <- list.files(merged_path, pattern = "_merged\\.tif$", full.names = TRUE) #(Already loaded above)

for (file in merged_files) {
  loop_start <- Sys.time()

  # Extract year from file name
  yr <- sub(".*_(\\d{4})_merged\\.tif", "\\1", basename(file))
  message("Processing vegetation index: ",VIndex," for year: ", yr, "  (file: ", basename(file), ")")

  # Load the merged raster (multi-band)
  year_raster <- rast(file)
  
  # extract bands as SpatRaster layers according to band_map
  # if a requested band index is beyond number of layers, throw error
  nlyr <- nlyr(year_raster)
  
  # helper to get band if exists, otherwise NULL
  get_band_or_null <- function(bandnum) {
    if (is.null(bandnum)) return(NULL)
    if (bandnum <= 0 || bandnum > nlyr) return(NULL)
    return(year_raster[[bandnum]])
  }
  
  bands <- list(
    blue  = get_band_or_null(band_map$blue),
    green = get_band_or_null(band_map$green),
    red   = get_band_or_null(band_map$red),
    nir   = get_band_or_null(band_map$nir),
    swir1 = get_band_or_null(band_map$swir1),
    swir2 = get_band_or_null(band_map$swir2)
  )
  
  # compute requested indices (this returns a named list of SpatRasters)
  VI_layers <- compute_indices_from_bands(bands, VIndex)
  
  # combine to a single SpatRaster with proper names and order
  indices <- rast(VI_layers)           # combine list into SpatRaster
  names(indices) <- names(VI_layers)   # ensure names are kept
  
  # Save NDVI raster
  out_file <- file.path(composites_folder, paste0(VIndex, "_" , yr, ".tif"))
  writeRaster(indices, filename = out_file, overwrite = TRUE)

  # Store file path for stacking later
  VIs_files <- c(VIs_files, out_file)

  # Clean up
  rm(year_raster, VI_layers, bands)
  terra::tmpFiles(current = TRUE, remove = TRUE)
  gc()

  loop_end <- Sys.time()
  message("Time for year ", yr, ": ", round(loop_end - loop_start, 1), " ", units(loop_end - loop_start))
}

# --- Build VI stack ---

VI_stack <- rast(VIs_files)

# Rename layers by year

VIs_files <- list.files("D:/Documents/QuebecTaigaStudy/Composites/NDVI/ExtraComposites/MaskedComposites", pattern = "\\.tif$", full.names = TRUE, recursive = TRUE)

VIyears <- as.integer(str_extract(basename(VIs_files), "\\d{4}"))
names(VI_stack) <- paste0(VIndex, "_", rep(years, each = length(VIndex)))

# # Save final stack
# writeRaster(VI_stack, file.path(composites_folder, "NDMI_timeseries_stack.tif"), overwrite = TRUE)
# VI_timeseries_stack <- rast(file.path(composites_folder, "NDMI_timeseries_stack.tif"))

```

### Compute spatial early warning signals

Set up and define the Input/Output paths, files and parameters.

```{r}
## ---- Define Input/Output paths and files ----

# Select vegetation index and resolution to process

VI <- 1  # 1-NDVI, 2-NDMI, 3-SAVI 4- LAI
RES <- 4 # 1-RES5, 2-RES6, 3-RES7, 4-RES8, 5-RES9
cores <- 8 # Number of cores to use for parallel processing

# Specify VI composite type (just for file naming convention)
# For example: "Raw", "Mean", "Median", "Quality" (Max = Quality)
composite_type <- "Max"

indices <- c("NDVI","NDMI","SAVI", "LAI")

index_name <- indices[VI]    # <<< change per job

composites_folder <- file.path("D:/Documents/QuebecTaigaStudy/Composites",
                               index_name,"Clean") #, "Clean", TwoYearComposites")

grids_folder <- "D:/Documents/QuebecTaigaStudy/Shapefiles/Grids/DGGRIDS/IGEO7"

grid_files <- list.files(grids_folder, pattern = paste0(".shp"), full.names = TRUE)
grid_resolutions <- c(5,6,7,8,9)
resolution <- grid_resolutions[RES]

grid_path <- paste0(grids_folder, "/IGEO7_", resolution, "RES.shp")   # Location of exported GEE composites
grid <- vect(grid_path) # grid object
grid$id <- 1:nrow(grid)
grid_sf <- sf::st_as_sf(grid)
#plot(grid)
# plot(grid_sf$geometry, col = "transparent", border = "blue")#, add= TRUE)

# Get scale from grid file name
scale <- as.numeric(gsub(".*_(\\d+)RES\\.shp$", "\\1", basename(grid_path)))

# Create VI-specfic output folders

out_folder <- "D:/Documents/QuebecTaigaStudy/SpatialAnalysis/IGEO7Results"    # Output folder for results
out_dir <- file.path(out_folder, index_name, file_path_sans_ext(basename(grid_path)))

if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

## Build file list for this index ----
raster_files <- list.files(
  composites_folder,
  pattern = paste0(index_name, "_\\d{4}\\.tif$"), ## Files are either named Composite_YYYY_merged ("_merged\\.tif$",) o INDEX_YYYY (paste0(index_name, "_\\d{4}\\.tif$"))
  full.names = TRUE
)

years <- as.integer(str_extract(basename(raster_files), "\\d{4}")) ### For files named Composite_YYYY_merged

cat("Processing", index_name, "rasters at level", scale, "resolution using", cores, "cores")
```

Functions to calculate the spatial EWS

```{r}
## Compute sews with NAs ####---
na_aware_ews <- function(mat, cg_subsize) {
  require(moments) # required packages
  require(raster)
  mat_coarse <- coarse_grain(mat, cg_subsize)
  c(skewness = skewness(as.vector(mat_coarse), na.rm = TRUE),
    variance = var(as.vector(mat_coarse), na.rm = TRUE),
    mean = mean(as.vector(mat_coarse), na.rm = TRUE),
    moran = Moran(raster(mat_coarse))) #Using the raster package (##This function calculates Queens case. The default setting uses a 3x3 neighborhood to compute "Queen's case" indices.)
    #moran = queen_moran(mat_coarse)) #Using sdep package (extremely time consuming for pixel level)
}
```

Parallel processing plan.

```{r}

## In each job’s plan(multisession, workers = N), choose N ≈ (total_cores / number_of_jobs)

future::plan(multisession, workers = cores)
# future::plan(sequential)

# Parallel loop: one worker per year
process_year  <- function(file, year, grid) {

  pid <- Sys.getpid()         # which worker handled this task
  task_start <- Sys.time()    # start time

  message("Year ", year, " started on worker PID ", pid, " at ", task_start)
  #loop_start <- Sys.time()
  #message("Starting processing for layer ", layer_index, " of ", length(raster_files), " at ", Sys.time())
  #cat("Processing year:", time_seq[layer_index], "\n")
  #message("Processing year: ", year)

  # Open raster inside worker
  current_layer <- rast(file) #raster_files[layer_index]
  indictestresults <- list()

  # Prepare results for this year
  results <- data.frame(
    id = grid$id,
    global_id = grid$global_id,
    year = year, #time_seq[layer_index]
    variance = NA_real_,
    skewness = NA_real_,
    moran = NA_real_,
    mean = NA_real_
  )

  # Loop over tiles (sequential within worker)
  for (i in 1:nrow(grid)) {
    current_tile <- grid[grid$id == i, , drop = FALSE]
    current_tile_vect <- terra::vect(current_tile)
    current_tile_vect <- terra::project(current_tile_vect, crs(current_layer))
    clipped_raster <- terra::crop(current_layer, current_tile_vect, mask = TRUE, touches = FALSE)
    clipped_raster <- raster(clipped_raster)
    raster_matrix  <- as.matrix(clipped_raster)

    if (all(is.na(raster_matrix))) next

    # --- Compute observed indicators ---
    indicators <- compute_indicator(raster_matrix, fun = na_aware_ews, cg_subsize = 1)

    indicatorvalues <- indicators$value

    results$variance[i] <- indicatorvalues["variance"]
    results$skewness[i] <- indicatorvalues["skewness"]
    results$moran[i]    <- indicatorvalues["moran"]
    results$mean[i]     <- indicatorvalues["mean"]
    
  }

  task_end <- Sys.time()
  runtime <- as.numeric(difftime(task_end, task_start, units = "secs"))

  message("Year ", year, " finished on worker PID ", pid, " at ", task_end,
          " (runtime: ", round(runtime, 2), "s)")

  # Add worker + runtime columns for diagnostics
  #results$worker <- pid
  #results$runtime <- runtime

  #loop_end <- Sys.time()
  #message("Finished processing for layer ", layer_index, " at ", Sys.time())
  #message("Time taken: ", round(loop_end - loop_start, 2), " secs")

  return(results)
}

## Parallel loop across years ----#
start_time <- Sys.time()

sews_results <- list()

sews_results <- future_lapply(
  seq_along(raster_files),
  function(idx) {
    process_year(raster_files[idx], years[idx], grid_sf)
  },
  future.packages = c(
    "terra", "raster", "spatialwarnings",
    "spatialwarningsGis", "sf", "dplyr"
  )
)

end_time <- Sys.time()

cat("Finished", index_name, "in",
    round(difftime(end_time, start_time, units = "mins"), 2),
    "minutes\n")
```

```{r}

### ---- 5. Save output ----

sews_results_df <- bind_rows(sews_results)

start_year <- min(years)
end_year <- max(years)

saveRDS(sews_results_df, file.path(out_dir, paste0("SEWS_IGEO7_", composite_type , index_name, "_" , start_year, "_", end_year, "_Summer0609_", scale , "RES", ".rds")))

write.csv(sews_results_df, file.path(out_dir, paste0("SEWS_IGEO7_", composite_type , index_name, "_" , start_year, "_", end_year, "_Summer0609_", scale, "RES", ".csv")), row.names = FALSE)

```

